---
title: "Homework 5"
author: "Alireza Rafiei"
date: "2/19/2023"
output: 
  html_document: default
  pdf_document: default
---

<style>div.soln { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}</style>

```{r setup, include=FALSE, echo=T}
knitr::opts_chunk$set(echo = F, message = F, warning = F, collapse = T)
library(tidyverse)
```


***
1. Using only `rnorm`, draw a sample of 1000 random numbers from $\chi^2_7$. Plot a histogram. **1 point**
```{r}
chi2_with_rnorm = function(n, df, verbose=0){
  x = replicate(n, sum(rnorm(df)^2))
  if (verbose == 0){
  hist(x, breaks = 20)
  }
  
  if (verbose==1) return(x)
}
chi2_with_rnorm(1000, 7)
```
***
2. Plot the log-likelihood of the sample assuming a $\chi^2_{df}$ distribution under candidate `df` values from 1 to 14. **1 point**
```{r}
# sample size
n = 1000
# sequence of candidate degrees of freedom
df_seq = 1:14          
# calculate log-likelihood for each degree of freedom value
log_lik = sapply(df_seq, function(df) sum(dchisq(chi2_with_rnorm(n, 7, verbose=1), df , log = TRUE)))
plot(df_seq, log_lik, type = "l", xlab = "Degrees of freedom", ylab = "Log-likelihood") 
```

***
3. Using only `rnorm` (and potentially your answer from question 1) draw a sample of 1000 random numbers from $t_7$. Plot a histogram. **1 point**
```{r}
t_with_rnorm = function(n, df, verbose=0){
  norm_sample = rnorm(1000)
  chi2_sample = chi2_with_rnorm(n, df, verbose=1)
  t_sample = norm_sample / sqrt(chi2_sample/df)
  
  if (verbose == 0){
  hist(t_sample, breaks = 20)
  }
  if (verbose==1) return(t_sample)
}
t_with_rnorm(1000, 7)
```
***
4. Plot the log-likelihood of your sample assuming a $t_{df}$ distribution under candidate `df` values from 1 to 14. **1 point**
```{r}
# sample size
n = 1000
# sequence of candidate degrees of freedom
df_seq = 1:14          
# calculate log-likelihood for each degree of freedom value
log_lik = sapply(df_seq, function(df) sum(log(dt(t_with_rnorm(n, 7, verbose=1), df))))
plot(df_seq, log_lik, type = "l", xlab = "Degrees of freedom", ylab = "Log-likelihood") 
```
***
5. Using only `rnorm` (or potentially your answer from question 1), draw a sample of 1000 random numbers from $F_{4,4}$. Plot a histogram. **1 point**. 
```{r}
F_with_rnorm = function(n, df1, df2, verbose=0){
  Z1 = chi2_with_rnorm(n, df1, verbose=1)
  Z2 = chi2_with_rnorm(n, df2, verbose=1)
  F_sample = (Z1/df1)/(Z2/df2)
  
  if (verbose == 0){
  hist(F_sample, breaks = 20)
  }
  if (verbose==1) return(F_sample)
}
F_with_rnorm(1000, 4, 4)
```
***
6. Plot the log-likelihood of your sample assuming a $F_{df_1,df_2}$ distribution with $df_1$ fixed at 4 under candidate `df_2` values from 1 to 14. **1 point**
```{r}
# sample size
n = 1000
# sequence of candidate degrees of freedom
df2_seq = 1:14          
# calculate log-likelihood for each degree of freedom value
log_lik = sapply(df2_seq, function(df2) sum(log(df(F_with_rnorm(1000, 4, 4, verbose = 1), df1 = 4, df2 = df2))))
plot(df2_seq, log_lik, type = "l", xlab = "Degrees of freedom", ylab = "Log-likelihood") 
```
***
7. Write a short function to convert a test statistic $x$ into the area under the $z$-density for values $≥|x|$. This area is the P value. The function should take $x$ as an argument. **(1 point)**
```{r}
p_value_z <- function(x) {
  pval <- 2 * (1 - pnorm(abs(x)))
  return(pval)
}

# Example
x = 1.5  
p_value_z(x)
```
***
8. Write a short function to convert a test statistic $x$ into the area under the $t_{df}$-density for values $≥|x|$. This area is the P value. The function should take $x$ and $df$ as arguments. **(1 point)**
```{r}
p_value_t = function(x, df) {

  tail_prob = pt(abs(x), df, lower.tail = FALSE)
  # In the case of a one-tailed hypothesis test, comment the following line.
  pval = 2 * tail_prob
  
  return(pval)
}

# Example
x = 1.3
df = 12
p_value_t(x, df)
```
***
9. Simulate the results of a cancer trial.

Let's say that in a double-blind randomized clinical breast cancer trial, the primary outcome is *tumor diameter*.
At study end, the average tumor reduction in tumor size was 3 ± 1 cm (mean ± sd) in the treatment group and 0 ± 1 cm (mean ± sd) in the control group.
Simulate reduction values for N = 12 treatment and N = 12 control participants. Use `t.test` to whether the reductions differ between groups.
Compare the results when using the default options and when using `var.equal = T`. **(1 point)**
```{r}
treatment <- rnorm(12, mean = 3, sd = 1)
control <- rnorm(12, mean = 0, sd = 1)

t_result1 <- t.test(treatment, control)
t_result2 <- t.test(treatment, control, var.equal = TRUE)

t_result1
t_result2
```
***
10. Recreate the functionality of `t.test`. Use base R functions (e.g., `length`, `mean`, `var`, `sqrt`) and `pt` to recreate the t statistic and p value from problem 9 without `t.test` using `var.equal=TRUE`. **(1 point)**
```{r}
# Simulate tumor diameter reduction values for treatment and control groups
treatment <- rnorm(12, mean = 3, sd = 1)
control <- rnorm(12, mean = 0, sd = 1)

# Calculate the sample means
mean_treatment <- mean(treatment)
mean_control <- mean(control)

# Calculate the sample variances
var_treatment <- var(treatment)
var_control <- var(control)

# Calculate the pooled standard deviation
pooled_sd <- sqrt(((length(treatment)-1)*var_treatment + (length(control)-1)*var_control) / (length(treatment) + length(control) - 2))

# Calculate the degrees of freedom
df <- length(treatment) + length(control) - 2

# Calculate the t statistic
t_statistic <- (mean_treatment - mean_control) / (pooled_sd * sqrt(1/length(treatment) + 1/length(control)))

# Calculate the P-value
p_value <- pt(abs(t_statistic), df = df, lower.tail = FALSE) * 2

# Print the results
cat("t statistic =", t_statistic, "\n")
cat("degrees of freedom =", df, "\n")
cat("P-value =", p_value, "\n")
```